In this dissertation, we have studied the likelihood-free inference
approaches, focusing on the novel ROMC method that we implemented in
an open-source software package. The main contribution of the
dissertation is this implementation which can be used mainly by the
research community for further experimentation.

In chapter 2, we presented the simulator-based models explaining the
particularities of the inference when the likelihood is not
tractable. We then presented an overview of the Optimisation
Monte-Carlo methods (OMC and ROMC) examining their strategies on
approximating the posterior. We discussed the point-of-view of the
ROMC approach, demonstrating the mathematic modelling it
introduces. We also presented the aspect of ROMC as a
meta-algorithm. Finally, at the end of chapter 2, we transformed the
mathematical modelling of ROMC into algorithmic form. Up to this
point, the dissertation mainly restates the ideas presented in the
original paper \autocite{Ikonomov2019} and, hence, it can be used
together with the paper by a reader who wants to understand the ROMC
approach.

The notable contribution of the dissertation is the implementation of
the method in the ELFI package. Due to the novelty of the method, it
has not been implemented in any other package so far. We tried to
implement the method focusing on four principles: simplicity,
accuracy, efficiency and extensibility. We tried to provide the user
with a simple-to-use method. Therefore, we followed the guidelines of
the ELFI package, aligning ROMC with the other provided inference methods. We
also kept the function calls simple, asking only for the necessary
arguments. Secondly, we tested the implementation on a range of
examples for ensuring accurate inference results. We designed some
artificial examples for evaluating our implementation using
ground-truth information. We also tested that it works smoothly under
general models, not artificially created by us. Thirdly, we tried to
solve the tasks efficiently. We applied parallel processing which offers a major speedup in the the execution time of all tasks. Furthermore, we avoided using redundant calls, we
exploited vectorisation when possible, and we avoided expensive
for-loops.  We measured the execution time of the method for providing
an overview of the time needed for performing each task. Finally, we
preserved extensibility; this was a significant priority in the
implementation design. Apart from offering a ready-to-use inference
method, we want our implementation to serve as the initial point for
researchers who would like to further experiment with the method. This
requirement aligns with the nature of ROMC as a meta-algorithm; one
can replace the method involved in a specific task without the rest of
the algorithm to collapse. Finally, we provide extensive documentation
of all implemented methods and a collection of examples for
illustrating the main use-cases. We wanted the reader to be able to
observe and interact in-practice with the functionalities we offer.